A White Noise Approach to Evolutionary Ecology

### 2020-04-03 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
mention lande 1975

To stay within the realm of biological plausibility we require a set technical assumptions. First, we assume the initial abundance density is continuous, non-negative, integrable and has finite trait mean and variance. That is, we assume $\nu(x,0)$ is continuous in $x$, satisfies $\nu(x,0)\geq0$ for all $x\in\mathbb{R}$ and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(\nu(x,t),x)$ that is differentiable with respect to both arguments and satisfies $m(y,x)\leq r$ for some $r\in\mathbb{R}$ and for all $x\in\mathbb{R}$, $y\geq0$. From here on we abbreviate $m(\nu(x,t),x)$ to $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities aside, the demographic dynamics can be modelled by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}
with the initial condition $\nu(x,0)$ described above. This PDE is semilinear due to the dependency of the growth rate $m(\nu,x)$ on the solution $\nu(x,t)$ and is referred to as a scalar reaction-diffusion equation [@lawrenceevans2010]. When $\mu=0$, equation (\ref{eq1}) can be seen as an analog of equation (\ref{first}) for a continuum of types. By assuming mutation acts as diffusion the effect of mutation causes $\nu(x,t)$ to flatten out over time. In fact, if the growth rate is constant across $x$, then this model of mutation will cause $\nu(x,t)$ to converge to a flat line as $t\to\infty$. Although clearly an idealized representation of biological reality, this model is sufficiently general to capture a large class of dynamics including density dependent growth and frequency dependent selection. As an example, logistic growth combined with quadratic stabilizing selection can be captured using the growth rate $m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\nu(x,t)$ where $a\geq0$ is the strength of stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c\geq0$ captures the effect of intraspecific competition and $r\in\mathbb{R}$ is the intrinsic growth rate in the absence of abiotic selection.

@@

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline}
Using our assumptions on mutation and rate of growth, we show in SM \S\ref{finite} $\nu(x,t)$ is twice differentiable with respect to $x$ and $\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty$ for all $t\geq0$. This implies that we are justified in swapping the order of differentiation and integration and the result $\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0$ can be derived from the fundamental theorem of calculus. Biological reasoning agrees with this latter result since mutation neither creates nor destroys individuals, but merely changes their type from their parental type. Taking the same approach, we derive the dynamics of the phenotypic distribution $p(x,t)$ in response to selection and muation as

\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}

This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which has been referred to as the Gaussian allelic approximation [in reference to the distribution of mutational effects at a given locus of a genome on the values of traits, @Lande1975; but see also @9780471986539], the infinitesimal model [in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect, @Barton2017] and the Gaussian descendants approximation [in reference to offspring trait values being normally distributed around their parental values, @Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of the continuum-of-alleles model.

@@

@@

However, it turns out that realizations of white noise do not exist as functions in the classical sense. Indeed, since Brownian motion is nowhere differentiable with respect to time, white noise cannot be formally understood as its time derivative. Thus our notation $\dot W$ is only meant to aid intuition and not be taken as formal. A formal understanding is possible by considering white noise as a _measure_-valued process [@Dawson1975; @Walsh] or as a _generalized_ process that acts on classically defined functions or stochastic processes to return either random variables or stochastic processes [@Krylov1981; @DaPrato2014]. Since a measure-valued process can be defined from a generalized process and a generalized process can be defined from a measure-valued process, the distinction between the two is more or less a matter of perspective. However, we find the perspective of white-noise as a generalized process to be a more efficient route for developing heuristics to help with some routine calculations involved with deriving SDE from SPDE. Hence, the notion of a generalized process provides the general idea implemented here. Although the treatments of SPDE provided by Krylov and Rozovskii [-@Krylov1981] and Da Prato and Zabczyk [-@DaPrato2014] extend the theory of SDE to formally treat SPDE in a general and elegant fashion, they require the navigation of an enourmous amount of technical definitions and detailed proofs. To extract some particularly useful results from this theory relevant to our goal of synthesizing the stochastic dynamics of biological populations, we provide a streamlined approach by capitolizing on the solid ground these authors have established. For instance, instead of rigorously proving properties of white-noise, we simply define them to be so, taking solice in the fact that the technical details have been worked out elsewhere. In SM \S\ref{DPZ14} we show how our informal treatment is related to the rigorous treatment provided by Da Prato and Zabczyk [-@DaPrato2014].

@@

To model imperfect heritability we consider the relationship between expressed phenotypes $x\in\mathbb{R}$ and associated genetic values $g\in\mathbb{R}$ known as _breeding values_. The breeding value of an individual is the sum of additive effects of the alleles carried by the individual on its expressed trait. Since our derivations of evolutionary equations are based on branching processes that assume asexually reproducing populations (\S\ref{stochastic}), the additive genetic variance $G$ is just the variance of breeding values in a population. For a detailed treatment of breeding values and additive genetic variances, see Bulmer [-@bulmer1980] and Lynch and Walsh [-@michaellynch1998]. Following standard assumptions of classical quantitative genetics we assume that the expressed trait for any given individual is normally distributed around their breeding value with variance $\eta$. Hence, $\sigma^2=G+\eta$. This coincides with assuming breeding values can be predicted from expressed traits using ordinary least squares. In the case that all of the effects of alleles on an expressed trait are additive, $\eta$ is known as the _variance of environmental deviation_ [@michaellynch1998]. For a given breeding value, we denote the probability density of a randomly drawn expressed trait by $\psi(x,g)$. Hence,
\psi(x,g)=\frac{1}{\sqrt{2\pi\eta}}\exp\left(-\frac{(x-g)^2}{2\eta}\right).
To include this relationship in our framework, we write $\rho(g,t)$ as the abundance density of breeding values at time $t$ so that $\int_{-\infty}^{+\infty}\rho(g,t)dg=\int_{-\infty}^{+\infty}\nu(x,t)dx=N(t)$. We switch our focus from directly modelling the evolution of $\nu(x,t)$ to modelling the evolution of $\rho(g,t)$. Once $\rho(g,t)$ is determined, we can compute $\nu(x,t)$ via
\begin{equation}\label{nu_from_rho}
\nu(x,t)=\int_{-\infty}^{+\infty}\rho(g,t)\psi(x,g)dg.
However, since selection acts on expressed phenotypes, we use our assumed relationship between breeding values and expressed traits to calculate the fitness of breeding values. To motivate our approach, consider the problem of inferring the breeding value of an individual given its expressed trait $x$. Denote $\mathfrak{g}$ a random variable representing the unkown breeding value. Under our model of inheritance we know $x$ is a random sample from a normal distribution with mean $\mathfrak{g}$ and variance $\eta$. Maximizing likelihood suggests $x$ is our best guess for $\mathfrak{g}$, but the actual value of $\mathfrak{g}$ is normally distributed around $x$ with the variance $\eta$. Hence, for fixed $x$, we obtain $\psi(x,g)$ as the probability density of $\mathfrak{g}$. Thus, the mean fitness of a breeding value $g$ across all individuals carrying $g$ can be written as
\begin{equation}\label{m_rel}
m^*(\rho,g)=\int_{-\infty}^{+\infty}m(\nu,x)\psi(x,g)dx.
This is similar to the approach taken by Kimura and Crow [-@Kimura1978] to calculate the overall effects of selection for expressed characters onto the changes in the distribution of alleles encoding those characters. However, instead of focusing on the frequencies of alleles at particular loci, we focus on the densities of breeding values. With the relationship between $m(\nu,x)$ and $m^*(\rho,g)$ established, we define the evolution of $\rho(g,t)$ by the SPDE
\dot\rho(g,t)=m^*(\rho,g)\rho(g,t)+\frac{\mu}{2}\frac{\partial^2}{\partial^2 g}\rho(g,t)+\sqrt{V\rho(g,t)}\dot W(g,t).
We assume $\rho(g,t)$ is Gaussian which implies its mode coincides with $\bar x$. Furthermore, since $\sigma^2=G+\eta$, we can use equation (\ref{m_rel}) and the chain rule from calculus to find (see SM \S\ref{fit2fit})
\frac{\partial\bar m}{\partial G}=\frac{\partial\bar m}{\partial\sigma^2}\frac{\partial\sigma^2}{\partial G}=\frac{\partial\bar m}{\partial\sigma^2},
\overline{\frac{\partial m}{\partial G}}=\overline{\frac{\partial m}{\partial\sigma^2}\frac{\partial\sigma^2}{\partial G}}=\overline{\frac{\partial m}{\partial\sigma^2}}.
Thus, equations (\ref{no_inher}) become
\begin{subequations}\label{inher}
\begin{equation}\label{xbarfinal}
d\bar x=G\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right)dt+\sqrt{V\frac{G}{N}}dW_2,
\begin{equation}\label{Gfinal}
dG=2G^2\left(\frac{\partial\bar m}{\partial G}-\overline{\frac{\partial m}{\partial G}}\right)dt+\left(\mu-V\frac{G}{N}\right)dt+ G\sqrt{\frac{2V}{N}}dW_3.

From expressions (\ref{inher}) we see that, under our model of inheritance, focusing on additive genetic variance $G$ instead the variance in expressed traits $\sigma^2$ makes no structural changes to the basic equations describing the dynamics of populations.

@@

A notable criticism of using niche overlap to measure the intensity of competition points to cases where populations competiting on multiple niche axes exhibit overlap on at least one of the axes, but no overall niche overlap [@Holt1987]. Thus niche overlap on lower-dimensional projections of some multivariate niche space does not imply the populations compete. To illustrate with a simple example, consider two populations competing for space on the plane $\mathbb{R}^2$. If the spatial distributions of the two populations overlap, then they will overlap on both spatial axes. However, if the populations do not overlap on at least one of the spatial axes, they will have no overall spatial overlap. Furthermore, even if the species overlap on both spatial axes, they need not have any overall spatial overlap. This final result corresponds to the fact that components of niche space do not necessarily interact multiplicatively to determine the consequences for the intensity of competition. In another component of Holt's [-@Holt1987] critique, an argument is made for the potential of competition occuring without any overlap in niche space. However, this argument is based on the practical difficulty of identifying every resource axis populations are competing on and how these axes interact to determine fitness consequences. Our model avoids these caveats by assuming competition only occurs along a single dimensional resource gradient. To map the degree of niche overlap to fitness, we assume competition between individuals with niche locations $x_i$ and $x_j$ additively decreases the Malthusian fitness for the individual in species $i$ by $c_i\mathcal{O}_{ij}(x_i,x_j)$ for some $c_i\geq0$. We refer to $c_i$ as the sensitivity to competition for species $i$. In SM \S\ref{diffuse} we combine this niche model with equations (\ref{N}), (\ref{xbarfinal}) and (\ref{Gfinal}) to find

@@

Temporal dynamics for each scenario are provided in Figure \ref{temporal}. This figure suggests weaker competition leads to smoother dynamics and a higher degree of organization within the community. Considering expression (\ref{N_comm}) we note that, all else equal, relaxed competition allows for larger growth rates which promote greater abundances. From (\ref{N_comm}) we also note that the per-capita effects on demographic stochasticity diminish with abundance. To see this, divide both sides by $N_i$. Inspecting expressions (\ref{x_comm}) and (\ref{G_comm}), we see that larger abundances also erode the effects of demographic stochasticity on the evolution of mean trait and additive genetic variance. These effects were already noted in \S\ref{equations}, and thus are not a consequence of our model of coevolution per-se, but we revisit them here since Figure \ref{temporal} demonstrates the importance of demographic stochasticity in structuring ecological communities even when populations are very large. Hence, contrary to the common assumption that stochastic effects can be ignored for large populations, we find that minute asymmetries by demographic stochasticity remain significant drivers of community structure. In particular, we initialized the species with identical state variables and model parameters, but found an enourmous amount of asymmetry and even some potential phase changes. In the following two paragraphs we describe the natural history of the community as illustrated in Figure \ref{temporal}.

After about $125.0$ units of time, the community appears to have shaken off the initial conditions and entered into a qualitatively distinct phase of dynamics. Aside from a few outliers, most of the species remain clustered together in their state variables. This lasts for approximately $375.0$ units of time until, at around time $500.0$, a drastic change occurs. At this moment the tightly packed cluster of species begins to fan out in all three state variables. Simultaneously, we observe large a shift in mean traits for higher values and in additive genetic variances for lower values. Upon inspecting our calculations, we diagnose the reason for this shift. The outlier species that were initially pushed away from the common abiotic optimum ($0.0$ in this case) evolved a significant reduction in the quantity of heritable variation ($\approx60\%$) due to directional selection induced by competition. This reduction in heritable variation slowed adaptation, causing these species to linger on the outskirts of niche space, some longer than others. In the meantime the rest of the community, being tightly packed, experienced greater competition which led to diminished abundances for these species and caused some members of the core group to veer away from the abiotic optimum. The reduced abundances of the core group led to reduced competition overall. As a result, the outlier populations were given a slight increase in growth rate, enough to allow them to increase their abundances orders of magnitude higher than the species in the core group and giving them more weight in driving the evolution of other species. Many of these heavy-hitting outlier species had already been maintaining negative mean traits, but around time $500.0$ the high abundance species with positive mean traits began to experience enough intraspecific competition to override interspecific competition and begin evolving towards the abiotic optimum. The sudden imbalance of these high abundance species effectively induced a single large competitive exclusion event pushing the majority of the community far away from the abiotic optimum. After this shift the cluster began to slowly bloom in all three state variables as species took advantage of novel asymmetries in their competitive abilities mediated by a new distribution of mean trait values across the community. About $125.0$ units of time later, the community reached a qualitatively new phase of dynamics. If we kept running the numerical integrator, we would continue to see similar drama unfolding over and over again as minute stochastic changes contribute to asymmetries which slowly build into drastic shifts.

@@

Relating our treatment of the niche to modern coexistence theory [@Chesson2000], the absolute competition coefficient $\alpha_{ij}$ becomes a dynamical quantity that can be written as

@@

Hence, our fitness function satisfies condition (\ref{m_cond}) of the main text.

>>>>

To stay within the realm of biological plausibility we require a set technical assumptions. First, we assume the initial abundance density is continuous, non-negative, integrable and has finite trait mean and variance. That is, we assume $\nu(x,0)$ is continuous in $x$, satisfies $\nu(x,0)\geq0$ for all $x\in\mathbb{R}$ and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(\nu(x,t),x)$ that is differentiable with respect to both arguments and satisfies $m(y,x)\leq r$ for some $r\in\mathbb{R}$ and for all $x\in\mathbb{R}$, $y\geq0$. From here on we abbreviate $m(\nu(x,t),x)$ to $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities aside, the demographic dynamics can be modelled by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}
with the initial condition $\nu(x,0)$ described above. This PDE is semilinear due to the dependency of the growth rate $m(\nu,x)$ on the solution $\nu(x,t)$ and is referred to as a scalar reaction-diffusion equation [@lawrenceevans2010]. Interpreting the trait value $x$ as spatial location, this model of population dynamics has been intensely studied [for a review, see @Cantrell2004]. When $\mu=0$, equation (\ref{eq1}) can be seen as an analog of equation (\ref{first}) for a continuum of types. By assuming mutation acts as diffusion the effect of mutation causes $\nu(x,t)$ to flatten out over time. In fact, if the growth rate is constant across $x$, then this model of mutation will cause $\nu(x,t)$ to converge to a flat line as $t\to\infty$. Although clearly an idealized representation of biological reality, this model is sufficiently general to capture a large class of dynamics including density dependent growth and frequency dependent selection. As an example, logistic growth combined with quadratic stabilizing selection can be captured using the growth rate 
\begin{equation}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\nu(x,t)
\end{equation}
where $a\geq0$ is the strength of stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c\geq0$ captures the sensitivity of fitness to intraspecific competition and $r\in\mathbb{R}$ is the intrinsic growth rate in the absence of abiotic selection. This exemplary fitness function has a few convenient properties that we breifly take note of. First, the effect of competition induces a carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrium abundance across all of $\mathbb{R}$. A stochastic generalization of this model, closely related to our model of diffuse coevolution presented later, has been studied by Dawson [-@Dawson1978].

@@

\begin{multline}\label{N_det}
\frac{d}{dt}N(t)=\frac{d}{dt}\int_{-\infty}^{+\infty}\nu(x,t)dx=\int_{-\infty}^{+\infty}\frac{\partial}{\partial t}\nu(x,t)dx \\
=\int_{-\infty}^{+\infty} m(\nu,x)\nu(x,t)dx+\int_{-\infty}^{+\infty}\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)dx \\
=N(t)\int_{-\infty}^{+\infty} m(\nu,x)p(x,t)dx=\bar m(t) N(t).
\end{multline}
Using our assumptions on mutation and rate of growth, we show in SM \S\ref{finite} $\nu(x,t)$ is twice differentiable with respect to $x$ and $\int_{-\infty}^{+\infty}\nu(x,t)dx<\infty$ for all $t\geq0$. This implies that we are justified in swapping the order of differentiation and integration and the result $\int_{-\infty}^{+\infty}\frac{\partial^2}{\partial x^2}\nu(x,t)dx=0$ can be derived from the fundamental theorem of calculus. Biological reasoning agrees with this latter result since mutation neither creates nor destroys individuals, but merely changes their type from their parental type. Taking the same approach, we derive the dynamics of the phenotypic distribution $p(x,t)$ in response to selection and muation as
\begin{multline}\label{p_PDE}
\frac{\partial}{\partial t}p(x,t)=\frac{\partial}{\partial t}\frac{\nu(x,t)}{N(t)}=\frac{1}{N^2(t)}\left(N(t)\frac{\partial}{\partial t}\nu(x,t)-\nu(x,t)\frac{d}{dt}N(t)\right) \\
=\frac{1}{N(t)}\Big(m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)-\bar m(t)\nu(x,t)\Big) \\
=(m(\nu,x)-\bar m(t)) p(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}p(x,t).
\end{multline}

This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which has been referred to as the Gaussian allelic approximation in reference to the distribution of mutational effects at loci in a genome on the values of traits [@Lande1975; but see also @9780471986539], the infinitesimal model in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect [@Barton2017] and the Gaussian descendants approximation in reference to offspring trait values being normally distributed around their parental values [@Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of Kimura's continuum-of-alleles model. We return to these concepts in \S\ref{inheritance} where we review a simple model of inheritance that leads us to an expression for the stochastic evolution of additive genetic variance.

@@

### Introduction to white noise

@@

However, it turns out that realizations of white noise do not exist as functions in the classical sense. Indeed, since Brownian motion is nowhere differentiable with respect to time, white noise cannot be formally understood as a time derivative. Thus our notation $\dot W$ is only meant to aid intuition and not be taken as formal. A formal understanding is possible by considering white noise as a _measure_-valued process [@Dawson1975; @Walsh] or as a _generalized_ process that acts on classically defined functions or stochastic processes to return either random variables or stochastic processes [@Krylov1981; @DaPrato2014]. Since a measure-valued process can be defined from a generalized process and a generalized process can be defined from a measure-valued process, the distinction between the two is more or less a matter of perspective. However, we find the perspective of white-noise as a generalized process to be a more efficient route for developing heuristics to help with some routine calculations involved with deriving SDE from SPDE. Hence, the notion of a generalized process provides the general idea implemented here. Although the treatments of SPDE provided by Krylov and Rozovskii [-@Krylov1981] and Da Prato and Zabczyk [-@DaPrato2014] extend the theory of SDE to formally treat SPDE in a general and elegant fashion, they require the navigation of an enourmous amount of technical definitions and detailed proofs. To extract some particularly useful results from this theory relevant to our goal of synthesizing the stochastic dynamics of biological populations, we provide a streamlined approach by capitolizing on the solid ground these authors have established. For instance, instead of rigorously proving properties of white-noise, we simply define them to be so, taking solice in the fact that the technical details have been worked out elsewhere. In SM \S\ref{DPZ14} we show how our informal treatment is related to the rigorous treatment provided by Da Prato and Zabczyk [-@DaPrato2014].

@@

To model imperfect heritability we consider the relationship between expressed phenotypes $x\in\mathbb{R}$ and associated genetic values $g\in\mathbb{R}$ known as _breeding values_. The breeding value of an individual is the sum of additive effects of the alleles carried by the individual on its expressed trait. Since our derivations of evolutionary equations are based on branching processes that assume asexually reproducing populations (\S\ref{stochastic}), the additive genetic variance $G$ is just the variance of breeding values in a population. For a detailed treatment of breeding values and additive genetic variances, see Bulmer [-@bulmer1980] and Lynch and Walsh [-@michaellynch1998].

Our treatment of the relationship between breeding values and expressed traits follows classical quantitative genetic assumptions such as those used in the seminal paper by Lande [-@Lande1975] to investigate the maintenance of genetic variation. In particular, we assume that the expressed trait for any given individual is independent of environmental conditions and normally distributed around their breeding value with variance $\eta$. Hence, $\sigma^2=G+\eta$. In the case that all of the effects of alleles on an expressed trait are additive, $\eta$ is known as the _variance of environmental deviation_ [@Lande1975; @michaellynch1998]. For a given breeding value, we denote the probability density of a randomly drawn expressed trait by $\psi(x,g)$. Hence,
\begin{equation}
\psi(x,g)=\frac{1}{\sqrt{2\pi\eta}}\exp\left(-\frac{(x-g)^2}{2\eta}\right).
\end{equation}
To include this relationship in our framework, we write $\rho(g,t)$ as the abundance density of breeding values at time $t$ so that $\int_{-\infty}^{+\infty}\rho(g,t)dg=\int_{-\infty}^{+\infty}\nu(x,t)dx=N(t)$. We switch our focus from directly modelling the evolution of $\nu(x,t)$ to modelling the evolution of $\rho(g,t)$. Once $\rho(g,t)$ is determined, we can compute $\nu(x,t)$ via
\begin{equation}\label{nu_from_rho}
\nu(x,t)=\int_{-\infty}^{+\infty}\rho(g,t)\psi(x,g)dg.
\end{equation}
However, since selection acts on expressed phenotypes, we use the assumed relationship between breeding values and expressed traits to calculate the fitness of breeding values. To motivate the approach taken, consider the problem of inferring the breeding value of an individual given its expressed trait $x$. Denote $\mathfrak{g}$ a random variable representing the unkown breeding value. Under this model of inheritance we know $x$ is a random sample from a normal distribution with mean $\mathfrak{g}$ and variance $\eta$. Maximizing likelihood suggests $x$ is our best guess for $\mathfrak{g}$, but the actual value of $\mathfrak{g}$ is normally distributed around $x$ with the variance $\eta$. Hence, for fixed $x$, we obtain $\psi(x,g)$ as the probability density of $\mathfrak{g}$. Thus, the mean fitness of a breeding value $g$ across all individuals carrying $g$ can be written as
\begin{equation}\label{m_rel}
m^*(\rho,g)=\int_{-\infty}^{+\infty}m(\nu,x)\psi(x,g)dx.
\end{equation}
This is similar to the approach taken by Kimura and Crow [-@Kimura1978] to calculate the overall effects of selection for expressed characters onto the changes in the distribution of alleles encoding those characters. However, instead of focusing on the frequencies of alleles at particular loci, our results focus on the densities of breeding values. With the relationship between $m(\nu,x)$ and $m^*(\rho,g)$ established, we define the evolution of $\rho(g,t)$ by the SPDE
\begin{equation}\label{rho_SPDE}
\dot\rho(g,t)=m^*(\rho,g)\rho(g,t)+\frac{\mu}{2}\frac{\partial^2}{\partial^2 g}\rho(g,t)+\sqrt{V\rho(g,t)}\dot W(g,t).
\end{equation}
Equation (\ref{rho_SPDE}) is a stochastic generalization of the deterministic PDE (\ref{eq1}) from \S\ref{deterministic}, but describes the evolution of the distribution of breeding values instead of expressed characters. In \S\ref{wnc} we review the origins of this equation and provide some theory to help make sense of it, particularly the term $\dot W$.

We assume $\rho(g,t)$ is Gaussian which implies its mode coincides with $\bar x$. Furthermore, since $\sigma^2=G+\eta$, we can use equation (\ref{m_rel}) and the chain rule from calculus (see SM \S\ref{fit2fit}) to find
\begin{subequations}
\begin{equation}
\frac{\partial\bar m}{\partial G}=\frac{\partial\bar m}{\partial\sigma^2}\frac{\partial\sigma^2}{\partial G}=\frac{\partial\bar m}{\partial\sigma^2},
\end{equation}
\begin{equation}
\overline{\frac{\partial m}{\partial G}}=\overline{\frac{\partial m}{\partial\sigma^2}\frac{\partial\sigma^2}{\partial G}}=\overline{\frac{\partial m}{\partial\sigma^2}}.
\end{equation}
\end{subequations}
Thus, equations (\ref{no_inher}) become
\begin{subequations}\label{inher}
\begin{equation}\label{xbarfinal}
d\bar x=G\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right)dt+\sqrt{V\frac{G}{N}}dW_2,
\end{equation}
\begin{equation}\label{Gfinal}
dG=2G^2\left(\frac{\partial\bar m}{\partial G}-\overline{\frac{\partial m}{\partial G}}\right)dt+\left(\mu-V\frac{G}{N}\right)dt+ G\sqrt{\frac{2V}{N}}dW_3.
\end{equation}
\end{subequations}

From expressions (\ref{inher}) we see that, under this model of inheritance, focusing on additive genetic variance $G$ instead the variance in expressed traits $\sigma^2$ makes no structural changes to the basic equations describing the dynamics of populations. In the next section, we make use of these expressions to develop a model of diffuse coevolution in a guild of $S$ species competing along a resource continuum.

@@

A notable criticism of using niche overlap to measure the intensity of competition points to cases where populations competiting on multiple niche axes exhibit overlap on at least one of the axes, but no overall niche overlap [@Holt1987]. Thus niche overlap on lower-dimensional projections of some multivariate niche space does not imply the populations compete. To illustrate with a simple example, consider two populations competing for space on the plane $\mathbb{R}^2$. If the spatial distributions of the two populations overlap, then they will overlap on both spatial axes. However, if the populations do not overlap on at least one of the spatial axes, they will have no overall spatial overlap. Furthermore, even if the species overlap on both spatial axes, they need not have any overall spatial overlap. This final result corresponds to the fact that components of niche space do not necessarily interact multiplicatively to determine the consequences for the intensity of competition. In another component of Holt's [-@Holt1987] critique, an argument is made for the potential of competition occuring without any overlap in niche space. However, this argument is based on the practical difficulty of identifying every resource axis populations are competing on and how these axes interact to determine fitness consequences. Our model avoids these caveats by assuming competition only occurs along a single dimensional resource gradient.

To map the degree of niche overlap to fitness, we assume competition between individuals with niche locations $x_i$ and $x_j$ additively decreases the Malthusian fitness for the individual in species $i$ by $c_i\mathcal{O}_{ij}(x_i,x_j)$ for some $c_i\geq0$. We refer to $c_i$ as the sensitivity to competition for species $i$. The term $c_i\mathcal{O}_{ij}(x_i,x_j)$ coincides with a special case of a term used to capture competition in Dawson's geostochastic logistic model, an SPDE model developed to study the combined effects of demographic stochasticity, spatial dispersion and locally finite carrying capacity [@Dawson1978]. In SM \S\ref{diffuse} we combine our treatment of resource competition with equations (\ref{N}), (\ref{xbarfinal}) and (\ref{Gfinal}) to find

@@

Temporal dynamics for each scenario are provided in Figure \ref{temporal}. This figure suggests weaker competition leads to smoother dynamics and a higher degree of organization within the community. Considering expression (\ref{N_comm}) we note that, all else equal, relaxed competition allows for larger growth rates which promote greater abundances. From (\ref{N_comm}) we also note that the per-capita effects on demographic stochasticity diminish with abundance. To see this, divide both sides by $N_i$. Inspecting expressions (\ref{x_comm}) and (\ref{G_comm}), we see that larger abundances also erode the effects of demographic stochasticity on the evolution of mean trait and additive genetic variance. These effects were already noted in \S\ref{equations}, and thus are not a consequence of our model of coevolution per-se, but we revisit them here since Figure \ref{temporal} demonstrates the importance of demographic stochasticity in structuring ecological communities even when populations are very large. Hence, contrary to the common assumption that stochastic effects can be ignored for large populations, we find that minute asymmetries generated by demographic stochasticity remain significant drivers of community structure. In particular, we initialized the species with identical state variables and model parameters, but found an enourmous amount of asymmetry and even some potential phase changes. In the following two paragraphs we describe the natural history of the community as illustrated in Figure \ref{temporal}.

We begin by describing the weak competition scenario. After about $125.0$ units of time, the community appears to have shaken off the initial conditions and entered into a qualitatively distinct phase of dynamics. Aside from a few outliers, most of the species remain clustered together in their state variables. This lasts for approximately $375.0$ units of time until, at around time $500.0$, a drastic change occurs. At this moment the tightly packed cluster of species begins to fan out in all three state variables. Simultaneously, we observe large a shift in mean traits for higher values and in additive genetic variances for lower values. Upon inspecting our calculations, we diagnose the reason for this shift. The outlier species that were initially pushed away from the common abiotic optimum ($0.0$ in this case) evolved a significant reduction in the amount of heritable variation ($\approx60\%$) due to directional selection induced by competition. This reduction in heritable variation slowed adaptation, causing these species to linger on the outskirts of niche space, some longer than others. In the meantime the rest of the community, being tightly packed, experienced greater competition which led to diminished abundances for these species and caused some members of the core group to veer away from the abiotic optimum. The reduced abundances of the core group led to reduced competition overall. As a result, the outlier populations were given a slight increase in growth rate, enough to allow them to increase their abundances orders of magnitude higher than the species in the core group and giving them more weight in driving the evolution of other species. Many of these heavy-hitting outlier species had already been maintaining negative mean traits, but around time $500.0$ the high abundance species with positive mean traits began to experience enough intraspecific competition to override interspecific competition. This generated a net selection gradient and associated evolutionary response towards the abiotic optimum. The sudden imbalance of these high abundance species effectively induced a single large competitive exclusion event pushing the majority of the community far away from the abiotic optimum. After this shift the cluster began to slowly bloom in all three state variables as species took advantage of novel asymmetries in their competitive abilities mediated by a new distribution of mean trait values across the community. About $125.0$ units of time later, the community reached a qualitatively new phase of dynamics. If we kept running the numerical integrator, we would continue to see similar drama unfolding over and over again as minute stochastic changes contribute to asymmetries which slowly build into drastic shifts.

@@

Relating our treatment of resource competition to modern coexistence theory [@Chesson2000], the absolute competition coefficient $\alpha_{ij}$ becomes a dynamical quantity that can be written as

@@

Hence, our fitness function is bounded above, as required in the main text.
